---
output: github_document
---

```{r}
#| label: knitr-opts
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# simevent

<!-- badges: start -->
[![CRAN
status](https://www.r-pkg.org/badges/version/simevent)](https://CRAN.R-project.org/package=simevent)
[![Codecov test
coverage](https://codecov.io/gh/BjarkeHautop/simevent/graph/badge.svg)](https://app.codecov.io/gh/BjarkeHautop/simevent)
<!-- badges: end -->

The `simevent` package provides tools for simulating and analyzing complex
continuous-time health care data. The simulated data includes variables that can
be interpreted as treatment decisions, disease progression, and health factors.

## Installation

You can install the stable version of simevent from CRAN with:

```r
install.packages("simevent")
```

or the development version of simevent from GitHub using pak:

```r
pak::pak("BjarkeHautop/simevent")
```

## Usage

```{r}
#| label: library

library(simevent)
```

The package builds event history data from a *graph*: baseline covariates, event
processes (censoring, terminal, or transient, with Weibull intensities), and the
Cox-type effects between them. `sim_graph()` defines the graph and
`sim_event_graph()` simulates from it. On top of this, the package provides:

- `sim_graph_from_fits()` for building a graph from `coxph()` fits to observed
  data, to simulate new data resembling it.
- `interval_format_data()` and `plot_event_data()` for reformatting and plotting
  simulated data.
- The `intervene` argument of `sim_event_graph()`, together with `event_risk()`,
  for simulating and evaluating interventions on processes and covariates.

A minimal example, simulating survival data for 100 individuals and plotting the
event histories:

```{r}
#| label: quick-example

graph <- sim_graph(
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1)
)
data <- sim_event_graph(graph, n = 100)
plot_event_data(data, title = "Survival Data")
```

For a full walkthrough of the simulation framework and all of the functions
above, see `vignette("sim-event-graph")`.

## Contributing

All contributions are welcome!
