---
output: github_document
---

```{r}
#| label: knitr-opts
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# simevent

<!-- badges: start -->
[![CRAN
status](https://www.r-pkg.org/badges/version/simevent)](https://CRAN.R-project.org/package=simevent)
[![Codecov test
coverage](https://codecov.io/gh/BjarkeHautop/simevent/graph/badge.svg)](https://app.codecov.io/gh/BjarkeHautop/simevent)
<!-- badges: end -->

The `simevent` package provides tools for simulating and analyzing complex
continuous-time health care data. The simulated data includes variables that can
be interpreted as treatment decisions, disease progression, and health factors.

## Installation

You can install the development version of `simevent` from
[GitHub](https://github.com/) with:

```r
# install.packages("pak")
pak::pak("miclukacova/simevent")
```

## Usage

```{r}
#| label: library

library(simevent)
```

At the core of the package is the flexible function `simEventData`, which
simulates event history data from a Cox proportional hazards model with Weibull
hazards. Users specify the number of events, and their intensity parameters and
covariate effects, to create custom scenarios. On top of this, the package
provides:

- Wrapper functions for common settings, e.g. `simSurvData` (survival data),
  `simCRdata` (competing risks), `simDisease`, `simTreatment`, `simDropIn`, and
  `simEventTV` (time-varying effects).
- Functions for simulating new data from a model already fitted to observed
  data: `simEventCox` (Cox models) and `simEventObj` (any model with a
  `predict2` method).
- `plotEventData` and `IntFormatData` for plotting and reformatting simulated
  data.
- `alphaSim` and `intEffectAlpha` for performing and evaluating interventions on
  the shape parameter of a process.

A minimal example, simulating survival data for 100 individuals and plotting the
event histories:

```{r}
#| label: quick-example

data <- simSurvData(100)
plotEventData(data, title = "Survival Data")
```

For a full walkthrough of the simulation framework and all of the functions
above, see `vignette("simevent")`.
