---
title: "Event Times and Time-Varying Markers"
vignette: >
  %\VignetteIndexEntry{Event Times and Time-Varying Markers}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-opts
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup
#| message: false

library(simevent)
library(survival)
library(ggplot2)
library(data.table)
```

This vignette shows two ways of making hazards depend on an individual's history
in `sim_model()`:

1. **Event times**, `T_<proc>.<k>`: effects that depend on *when* an event
   happened.
2. **Markers**, `sim_mark()`: covariates that are re-measured over time, e.g. a
   lab value drawn at each visit.

# Event Times: `T_<proc>.<k>`

Inside a `sim_effect()` expression, `T_<proc>.<k>` is the time of process
`proc`'s `k`-th event, and `0` if it hasn't happened yet. Combined with the
current time `t`, this gives effects on time since an event: `t - T_operation.1`
is the time since the first operation.

Since `T_<proc>.<k>` is `0` before the event, `t - T_operation.1` is just `t`
for someone not yet operated. Multiplying by the event count, as in
`operation * f(t - T_operation.1)`, can be used to have the effect off until the
event has happened.

## Example: Post-Operative Mortality

Suppose patients may undergo an operation, which raises the death hazard
immediately after, and then gradually fades off. Such an example could look like
this:

$$
  \lambda_{\text{death}}(t)
  = \lambda_0(t) \exp\left( 3 \cdot \text{operation}(t) \cdot e^{-2(t-T_{\text{operation},1})} \right).
$$

Here all baseline hazards are constant ($\nu = 1$), so $\lambda_0(t) = 0.02$.
After the operation, measured in time since the operation,
$u = t - T_{\text{operation},1}$, the death hazard is therefore

$$
  h(u) = 0.02 \exp\left( 3 e^{-2u} \right).
$$

Right after the operation the hazard is $h(0) = 0.02 e^3 \approx 0.40$, about 20
times the baseline. As time passes, $e^{-2u}$ goes to 0 and the hazard returns
to the baseline of $0.02$. Two time units after the operation, it is only about
6% above the baseline.

```{r}
#| label: operation-model

operation_model <- sim_model(
  censoring = sim_process("censoring", eta = 0.05, nu = 1),
  operation = sim_process("transient", eta = 0.2, nu = 1, limit = 1),
  death = sim_process("terminal", eta = 0.02, nu = 1),
  effects = list(
    sim_effect("operation * exp(-2 * (t - T_operation.1))", "death", coef = 3)
  )
)

operation_data <- sim_events(
  operation_model,
  n = 5000,
  max_cens = 10,
  seed = 1
)
head(operation_data)
```

To check the simulation, follow each operated patient from their operation until
death or censoring. The Nelson--Aalen estimator then estimates the cumulative
death hazard since the operation, which is (substituting $w = 3 e^{-2v}$, so
$dv = -dw / (2w)$)

$$
  \begin{aligned}
    H(u) & = \int_0^u h(v) \, dv
    = 0.02 \int_{3 e^{-2u}}^{3} \frac{e^w}{2w} \, dw \\
         & = 0.01 \left[ \operatorname{Ei}(3) - \operatorname{Ei}\left( 3 e^{-2u} \right) \right],
  \end{aligned}
$$

where $\operatorname{Ei}(x) = \int_{-\infty}^x e^w / w \, dw$ is the exponential
integral. Censoring and the end of follow-up don't depend on time since the
operation, so the estimate is unbiased. Below, $H(u)$ is computed by numerical
integration.

```{r}
#| label: operation-plot

operation_time <- operation_data[event == "operation", .(id, T_op = time)]
post_op <- merge(
  operation_data[, .SD[.N], by = id],
  operation_time,
  by = "id"
)
post_op[, u := time - T_op]

fit <- survfit(Surv(u, event == "death") ~ 1, data = post_op)

H <- function(u) {
  vapply(
    u,
    function(x) integrate(function(v) 0.02 * exp(3 * exp(-2 * v)), 0, x)$value,
    numeric(1)
  )
}
u_grid <- seq(0, 8, length.out = 200)

ggplot() +
  geom_step(aes(fit$time, fit$cumhaz, colour = "Nelson–Aalen")) +
  geom_line(aes(u_grid, H(u_grid), colour = "True H(u)"), linetype = "dashed") +
  coord_cartesian(xlim = c(0, 8)) +
  labs(
    x = "Time since operation",
    y = "Cumulative death hazard",
    colour = NULL
  )
```

Other shapes follow the same pattern. The `from` expression times `coef` is
added to the log-hazard, for example:

  | `from`                                | Effect on the log-hazard                          |
  | ------------------------------------- | ------------------------------------------------- |
  | `operation * (t - T_operation.1)`     | Grows linearly with time since the operation      |
  | `operation * (t - T_operation.1 < 1)` | Constant for one time unit after, then gone       |
  | `(relapse >= 2) * (t - T_relapse.2)`  | Grows linearly with time since the second relapse |

# Time-Varying Markers: `sim_mark()`

A `sim_covariate()` is drawn once at baseline and never changes. A `sim_mark()`
is drawn at baseline and then *redrawn* each time one of its `update` processes
fires, staying constant in between. It takes:

- `init`: a function of `N` (and earlier covariates) giving the baseline values,
  as for `sim_covariate()`.
- `update`: the transient processes whose events redraw the marker.
- `draw`: a function of `N` and `lp` giving the new values.

What a new value depends on is specified with `sim_effect()`s *into* the marker.
At each update, their `coef * from` terms are summed into the linear predictor
`lp`, which `draw` can then use, e.g. as the mean of `rnorm()`. These effects
can use anything a hazard effect can, including the marker's own value before
the update, and see the triggering event as having happened.

The marker can also be used in `sim_effect()` like any other covariate, and the
hazards always use its current value.

## Example: Blood Pressure and Stroke

Suppose patients have blood pressure measured at visits. High blood pressure
makes treatment initiation more likely, treatment lowers subsequent blood
pressure, and both blood pressure and treatment affect the stroke hazard. Blood
pressure is therefore a time-varying confounder of the effect of treatment on
stroke.

Let's define such a setup, where each new measurement depends on the previous
one and on treatment:

$$
  \text{bp}_{\text{new}}
  \sim N\left( 130 + 0.8 (\text{bp}_{\text{prev}} - 130) - 2 \cdot \text{treatment}, \; 6^2 \right).
$$

Untreated, blood pressure has mean 130. Treatment lowers each new measurement
mean by 2, and since each measurement carries over 0.8 of the previous one's
deviation, the effect converges to a mean of $130 - 2 / (1 - 0.8) = 120$.

```{r}
#| label: bp-model

bp_model <- sim_model(
  age = sim_covariate(function(N) rnorm(N, mean = 60, sd = 10)),
  visit = sim_process("transient", eta = 1, nu = 1),
  bp = sim_mark(
    init = function(N, age) rnorm(N, 130 + 0.5 * (age - 60), 12),
    update = "visit",
    draw = function(N, lp) rnorm(N, mean = 130 + lp, sd = 6)
  ),
  treatment = sim_process("transient", eta = 0.05, nu = 1, limit = 1),
  censoring = sim_process("censoring", eta = 0.05, nu = 1),
  stroke = sim_process("terminal", eta = 0.02, nu = 1.2),
  effects = list(
    sim_effect("bp - 130", "bp", coef = 0.8),
    sim_effect("treatment", "bp", coef = -2),
    sim_effect("bp - 130", "treatment", coef = 0.08),
    sim_effect("bp - 130", "stroke", coef = 0.04),
    sim_effect("treatment", "stroke", coef = -0.5)
  )
)
summary(bp_model)

bp_data <- sim_events(bp_model, n = 3000, max_cens = 10, seed = 1)
head(bp_data, 8)
```

The summary unrolls the effects into the marker over its updates: `bp[k+1]`, the
value after the `k+1`-th visit, is drawn from `bp[k]` and the current treatment
status. In the data, `bp` is the value after each row's event, and `bp_0` the
baseline value.

Let's visualize the blood pressure for a few patients, with the treatment start
marked:

```{r}
#| label: bp-plot

ids <- 1:6
baseline <- bp_data[id %in% ids, .(time = 0, bp = bp_0[1]), by = id]
trajectories <- rbind(baseline, bp_data[id %in% ids, .(id, time, bp)])
treated <- bp_data[id %in% ids & event == "treatment"]

ggplot(trajectories, aes(time, bp)) +
  geom_step() +
  geom_vline(
    aes(xintercept = time),
    data = treated,
    linetype = "dashed",
    colour = "red"
  ) +
  facet_wrap(~id, labeller = label_both) +
  labs(x = "Time", y = "Blood pressure")
```

## Recovering the Effects

To check the simulation works correctly we fit a Cox model for stroke, with each
row as an interval from the previous event to this one. Each row stores the
values *after* its event, so the values during the interval are those of the
previous row, or the baseline values for a patient's first row.
`interval_format_data()` does this shift, for marker columns (`mark_cols`) and
process counts (`proc_cols`):

```{r}
#| label: bp-cox

bp_int <- interval_format_data(
  bp_data,
  proc_cols = "treatment",
  mark_cols = "bp"
)

coxph(
  Surv(tstart, tstop, event == "stroke") ~ I(bp - 130) + treatment,
  data = bp_int
)
```

The estimates are close to the simulated effects of `0.04` and `-0.5`.

The effects into the marker can be checked the same way. On each `visit` row,
`bp` in `bp_int` is the previous measurement, and `bp` in `bp_data` the new one,
so regressing one on the other and treatment should give roughly `0.8`, `-2`,
and a residual standard deviation of `6`:

```{r}
#| label: bp-lm

visits <- bp_int[event == "visit"]
visits[bp_data, bp_new := i.bp, on = .(id, time)]

bp_fit <- lm(I(bp_new - 130) ~ 0 + I(bp - 130) + treatment, data = visits)
coef(bp_fit)
sigma(bp_fit)
```
