---
title: "Migrating to sim_events()"
vignette: >
  %\VignetteIndexEntry{Migrating to sim_events()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-opts
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup

library(simevent)
library(survival)

compare_model <- function(wrapper_data, model_data) {
  wrapper_data <- as.data.frame(wrapper_data)
  model_data <- as.data.frame(model_data)
  # The wrappers code events as integers; the models below declare their
  # processes in the same order as those codes.
  model_data$Delta <- as.integer(model_data$event) - 1L
  names(model_data)[match(c("id", "time"), names(model_data))] <-
    c("ID", "Time")
  isTRUE(all.equal(
    wrapper_data,
    model_data[names(wrapper_data)],
    tolerance = 1e-6
  ))
}
```

`compare_model()` checks that a `sim_events()` reproduction draws the same
random numbers, in the same order, as the wrapper it reproduces. Only holds
below for `simSurvData()`/`simCRdata()`.

`simDisease()`/`simTreatment()` have a `"transient"` process, so individuals
finish at different times: `sim_events()` only draws for whoever's still
running, while the wrapper draws for everyone and discards the unused draws for
those already finished.

# Introduction

Introduces the `sim_events()` API, a front end to the same simulator as
`sim.generic()`.

# The `sim_model()` Interface

A `sim_model()` call has three kinds of possible inputs:

- `sim_covariate()`: a baseline covariate generator. Takes `N` and, optionally,
  any other covariate defined earlier in the same call (matched by argument
  name).
- `sim_process()`: a `"censoring"`, `"terminal"`, or `"transient"` (fires at
  most `limit` times, default `Inf`) process, with Weibull `eta`/`nu` parameters
  or a given cumulative baseline hazard curve (`cumhaz`).
- `sim_effect()`: a `from -> to` edge with a Cox-type `coef`, where `from` is a
  covariate or process name and `to` is a process name.

```{r}
#| label: sim-model-basic

model <- sim_model(
  L0 = sim_covariate(function(N) rbinom(N, 1, 0.4)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1),
  effects = list(sim_effect("L0", "death", coef = 1.5))
)
model

data <- sim_events(model, n = 100)
head(data)
```

`sim_events()`'s `intervene` argument replaces `sim.generic()`'s
`alpha.intervention`/`baseline.intervention` pair with a single named list: give
a process name to scale its `eta`, or a covariate name to fix it to a constant.

```{r}
#| label: sim-model-intervention

data_intervened <- sim_events(
  model,
  n = 100,
  intervene = list(death = 0.2, L0 = 1)
)
head(data_intervened)
```

## Non-Linear and History-Dependent Effects

`sim_effect()`'s `from` doesn't have to name a single covariate or process. Any
other string is parsed and evaluated as an R expression, against every
covariate/process value plus the current time `t`, the event times
`T_<proc>.<k>` (time of `proc`'s `k`-th event, `0` if it hasn't happened yet)
and `last_time(proc)`. Covers thresholds, interactions, and nonlinear transforms
inline, and effects depending on a process's full occurrence history, not just
its current count, without a separate `sim_derived()` covariate.

For example, an effect that only fires on a process's 3rd occurrence, not the
1st or 2nd:

```{r}
#| label: sim-model-nonlinear-effect

relapse_model <- sim_model(
  censoring = sim_process("censoring", eta = 0.01, nu = 1),
  relapse = sim_process("transient", eta = 0.3, nu = 1, limit = 5),
  death = sim_process("terminal", eta = 0.01, nu = 1),
  effects = list(
    sim_effect("relapse == 3", "death", coef = 8)
  )
)

set.seed(1405)
relapse_data <- sim_events(relapse_model, n = 2000, max_events = 40)

deaths <- relapse_data[relapse_data$event == "death", ]
mean(deaths$relapse == 3)
```

Most deaths happen right at the 3rd relapse, not before or after.

`T_<proc>.<k>` and `last_time()` give access to timing, not just counts.
`T_<proc>.<k>` is the time of `proc`'s `k`-th event, and `0` if it hasn't
happened yet. Effects on time since an event therefore take the form
`proc * f(t - T_proc.k)`: multiplying by the event count switches the effect off
until the event has happened, whatever `f` is.

Here the death hazard spikes right after an operation and then fades:

```{r}
#| label: sim-model-event-time-effect

operation_model <- sim_model(
  censoring = sim_process("censoring", eta = 0.05, nu = 1),
  operation = sim_process("transient", eta = 0.2, nu = 1, limit = 1),
  death = sim_process("terminal", eta = 0.02, nu = 1),
  effects = list(
    sim_effect("operation * exp(-(t - T_operation.1))", "death", coef = 3)
  )
)

operation_data <- sim_events(operation_model, n = 2000)
head(operation_data)
```

Without the `operation *`, the not-yet-operated would have
`exp(-(t - 0)) = exp(-t)`, a spike at time 0 that never happened. For a risk
that instead builds up after the operation, use
`operation * (t - T_operation.1)`.

And here the death hazard drops for one time unit right after each checkup:

```{r}
#| label: sim-model-history-effect

checkup_model <- sim_model(
  censoring = sim_process("censoring", eta = 0.05, nu = 1.1),
  checkup = sim_process("transient", eta = 0.3, nu = 1),
  death = sim_process("terminal", eta = 0.05, nu = 1.1),
  effects = list(
    sim_effect("t - last_time(checkup) < 1", "death", coef = -2)
  )
)

checkup_data <- sim_events(checkup_model, n = 2000)
head(checkup_data)
```

`last_time(proc)` is `-Inf` before `proc`'s first occurrence, so
`t - last_time(proc) < 1` needs no `NA` handling.

# Reproducing the Preset Wrappers

## `simSurvData()`

`simSurvData()`'s default (`beta = NULL`, i.e. no covariate effects) is two
terminal processes, censoring and death, both with `eta = 0.1`, `nu = 1.1`:

```{r}
#| label: simsurvdata-equivalence

set.seed(1405)
wrapper_data <- simSurvData(200)

set.seed(1405)
model <- sim_model(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1)
)
model_data <- sim_events(model, n = 200)

compare_model(wrapper_data, model_data)
```

## `simCRdata()`

`simCRdata()` adds a third terminal process (a second competing cause):

```{r}
#| label: simcrdata-equivalence

set.seed(1405)
beta <- matrix(c(0.5, -1, -0.5, 0.5, 0, 0.5), ncol = 3, nrow = 2)
wrapper_data <- simCRdata(N = 200, beta = beta)

set.seed(1405)
model <- sim_model(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  cause1 = sim_process("terminal", eta = 0.1, nu = 1.1),
  cause2 = sim_process("terminal", eta = 0.1, nu = 1.1),
  effects = list(
    sim_effect("L0", "censoring", 0.5),
    sim_effect("A0", "censoring", -1),
    sim_effect("L0", "cause1", -0.5),
    sim_effect("A0", "cause1", 0.5),
    sim_effect("A0", "cause2", 0.5)
  )
)
model_data <- sim_events(model, n = 200)

compare_model(wrapper_data, model_data)
```

## `simDisease()`

`simDisease()` has a censoring process, a terminal death process, and a
covariate process `L` that can happen at most once and can itself affect death;
a `"transient"` process with `limit = 1`. Since `L` can fire, the simulation
loop can run more than once per individual, so `sim_events()`'s random draws no
longer line up one-to-one with `simDisease()`'s (see the note on
`compare_model()` above); event-type proportions are compared instead:

```{r}
#| label: simdisease-equivalence

set.seed(1405)
wrapper_data <- simDisease(
  N = 5000,
  cens = 1,
  eta = c(0.1, 0.3, 0.1),
  nu = c(1.1, 1.3, 1.1),
  beta_L0_L = 1,
  beta_A0_L = -1.1,
  beta_L_D = 1,
  beta_L0_D = 0
)

model <- sim_model(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.3, nu = 1.3),
  L = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
  effects = list(
    sim_effect("L0", "L", 1),
    sim_effect("A0", "L", -1.1),
    sim_effect("L", "death", 1)
  )
)
model_data <- sim_events(model, n = 5000)

rbind(
  simDisease = prop.table(table(wrapper_data$Delta)),
  sim_model = prop.table(table(model_data$event))
)
```

## `simTreatment()`

`simTreatment()` (with `op = 1`, the default) adds a second `"transient"`
process, treatment (`A`), which can itself affect and be affected by the
covariate process `L`. `simTreatment()` doesn't declare its own `A0`, so its
reproduction skips it too: `sim_model()` has no default covariates.

```{r}
#| label: simtreatment-equivalence

set.seed(1405)
wrapper_data <- simTreatment(
  N = 5000,
  beta_L_A = 1,
  beta_L_D = 1,
  beta_A_D = -1,
  beta_A_L = -0.5,
  beta_L0_A = 1,
  eta = rep(0.1, 4),
  nu = rep(1.1, 4),
  cens = 1,
  op = 1
)

model <- sim_model(
  L0 = sim_covariate(function(N) runif(N)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1),
  A = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
  L = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
  effects = list(
    sim_effect("L0", "death", 1),
    sim_effect("L0", "A", 1),
    sim_effect("L0", "L", 1),
    sim_effect("A", "death", -1),
    sim_effect("A", "L", -0.5),
    sim_effect("L", "death", 1),
    sim_effect("L", "A", 1)
  )
)
model_data <- sim_events(model, n = 5000)

rbind(
  simTreatment = prop.table(table(wrapper_data$Delta)),
  sim_model = prop.table(table(model_data$event))
)
```

## `simStatinData()`

`simStatinData()` mainly demonstrates non-default baseline covariate generators.
`sim_covariate()` reproduces this directly, with no forced covariate names to
work around:

```{r}
#| label: simstatindata-equivalence

set.seed(1405)
model <- sim_model(
  L0 = sim_covariate(function(N) rbinom(N, 1, 0.4)),
  A0 = sim_covariate(function(N, L0) pmin(rexp(N, 0.3) + 70, 100)),
  censoring = sim_process("censoring", eta = 0.025, nu = 1.1),
  death = sim_process("terminal", eta = 0.025, nu = 1.1)
)
model_data <- sim_events(model, n = 200)
summary(model_data$A0)
```

# Estimating Intervention Effects

`alphaSim()` and `intEffectAlpha()` simulate one of the preset settings with the
shape parameter `eta` of one process multiplied by `alpha`, and summarise the
result as the proportion of individuals experiencing death and the intervened
process by time $\tau$ (or the years lost to them before $\tau$). With the model
API this splits into two general steps, which work for any process in any model:

1. Simulate under the intervention with `sim_events()`'s `intervene` argument:
   `intervene = list(<process> = alpha)` multiplies that process's `eta` by
   `alpha`.
2. Summarise with `event_risk()`: the risk $P(T \le \tau)$ of each process's
   first event, or (`type = "time_lost"`) the expected time lost to it before
   $\tau$, $E[\tau - \min(T, \tau)]$.

Take `alphaSim()`'s `"Disease"` setting with `alpha = 0.5`, which halves the
hazard of the disease process `L`:

```{r}
#| label: alphasim-old

set.seed(1405)
alphaSim(
  N = 5000,
  eta = rep(0.1, 3),
  nu = rep(1.1, 3),
  alpha = 0.5,
  tau = 5,
  setting = "Disease"
)
```

The same model with `sim_model()`:

```{r}
#| label: intervention-model

disease_model <- sim_model(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1),
  L = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
  effects = list(
    sim_effect("L0", "death", 1),
    sim_effect("L0", "L", 1),
    sim_effect("L", "death", 1)
  )
)
```

`alphaSim()` switches censoring off by default (`cens = 0`). Simulating both
with and without the intervention gives the effect directly:

```{r}
#| label: intervention-risk

set.seed(1405)
observed <- sim_events(disease_model, n = 5000, cens = 0)
intervened <- sim_events(
  disease_model,
  n = 5000,
  cens = 0,
  intervene = list(L = 0.5)
)

risk_observed <- event_risk(observed, disease_model, c("death", "L"), tau = 5)
risk_intervened <- event_risk(
  intervened,
  disease_model,
  c("death", "L"),
  tau = 5
)
cbind(
  risk_observed[, "process"],
  observed = risk_observed$risk,
  intervened = risk_intervened$risk
)
```

Halving the disease hazard lowers the risk of disease, and, since disease raises
the death hazard (`sim_effect("L", "death", 1)`), the risk of death too.

`alphaSim()`/`intEffectAlpha()`'s `years_lost = TRUE` corresponds to
`type = "time_lost"`, and their `a0` argument (summarise only individuals with
`A0 == a0`) to `by = "A0"`, which summarises every group at once:

```{r}
#| label: intervention-time-lost

event_risk(
  intervened,
  disease_model,
  c("death", "L"),
  tau = 5,
  type = "time_lost",
  by = "A0"
)
```

`by` conditions on the observed value of `A0`. For a `do()`-style contrast,
where everyone's `A0` is set to a value, fix it with `intervene` instead, e.g.
`intervene = list(L = 0.5, A0 = 1)`.

Finally, `intEffectAlpha()`'s `plot = TRUE` is `plot_event_data()` on the
intervened data:

```{r}
#| label: intervention-plot

plot_event_data(
  intervened[intervened$id <= 100, ],
  title = "Under Intervention"
)
```

## Calibrating a Coefficient to a Target Effect

Hitting a specific target effect by hand means repeating the comparison above
many times: a 1-D search. Risk is monotonic in a Cox-type coefficient, so
`uniroot()` finds it directly from a function returning the simulated effect for
a given coefficient.

Use the same `n`/`tau` on every call, and the same random seed for the natural
and intervened runs (common random numbers), so only the coefficient and the
intervention vary, not independent Monte Carlo noise too. Otherwise the search
target is too noisy for `uniroot()` to bracket a root.

Take `disease_model` from above, but suppose disease's effect on death (fixed at
`1` there) is unknown, and the target is: halving the disease hazard should
lower 5-year death risk by 5 percentage points.

```{r}
#| label: calibration-objective

disease_effect_on_death <- function(coef, n = 3000, tau = 5, seed = 1) {
  model <- sim_model(
    L0 = sim_covariate(function(N) runif(N)),
    A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
    censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
    death = sim_process("terminal", eta = 0.1, nu = 1.1),
    L = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
    effects = list(
      sim_effect("L0", "death", 1),
      sim_effect("L0", "L", 1),
      sim_effect("L", "death", coef)
    )
  )

  set.seed(seed)
  natural <- sim_events(model, n = n, cens = 0)
  set.seed(seed)
  intervened <- sim_events(
    model,
    n = n,
    cens = 0,
    intervene = list(L = 0.5)
  )

  risk_natural <- event_risk(natural, model, "death", tau = tau)$risk
  risk_intervened <- event_risk(intervened, model, "death", tau = tau)$risk
  risk_natural - risk_intervened
}
```

Larger coefficients make disease matter more for death, so halving its hazard
lowers death risk by more: `disease_effect_on_death()` is increasing in `coef`.

```{r}
#| label: calibration-check-monotone

sapply(c(0, 1, 2, 3), disease_effect_on_death)
```

`uniroot()` finds the coefficient giving a 5 percentage point reduction:

```{r}
#| label: calibration-uniroot

calibrated <- uniroot(
  function(coef) disease_effect_on_death(coef) - 0.05,
  interval = c(0, 3),
  tol = 0.01
)
calibrated$root
```

For several coefficients and targets at once, wrap the sum of squared deviations
into one objective (still with common random numbers) and use `optim()` instead
of `uniroot()`.

# Building a `sim_model()` from Fitted Cox Models

`sim_model_from_fits()` is the model-native replacement for `sim.from.data()`:
instead of user-specified effects, it builds a `sim_model()` from a set of
fitted `coxph()` models (one per process) and the data they were fit to, so
simulated data mimics an observed dataset's distribution.

Unlike `sim.from.data()`, which requires building a `sim.parameters` list,
including a `baseline.summary` computed by hand from `mean()`/
`sd()`/`min()`/`max()`, `sim_model_from_fits()` resamples whole rows of baseline
covariates from the data itself, keeping each covariate's distribution and the
relationships between them.

```{r}
#| label: sim-model-from-fits-setup

# Some "observed" data, from a 3-cause competing-risks sim_model():
set.seed(1405)
observed_model <- sim_model(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  cause1 = sim_process("terminal", eta = 0.1, nu = 1.1),
  cause2 = sim_process("terminal", eta = 0.1, nu = 1.1),
  effects = list(
    sim_effect("L0", "censoring", 0.5),
    sim_effect("A0", "censoring", -1),
    sim_effect("L0", "cause1", -0.5),
    sim_effect("A0", "cause1", 0.5),
    sim_effect("A0", "cause2", 0.5)
  )
)
observed_data <- sim_events(observed_model, n = 1000)

# Refit from "observed" data, then rebuild a sim_model(), as if
# observed_model were unknown:
fits <- list(
  censoring = coxph(
    Surv(time, event == "censoring") ~ L0 + A0,
    data = observed_data
  ),
  cause1 = coxph(Surv(time, event == "cause1") ~ L0 + A0, data = observed_data),
  cause2 = coxph(Surv(time, event == "cause2") ~ L0 + A0, data = observed_data)
)
types <- c(censoring = "censoring", cause1 = "terminal", cause2 = "terminal")

model <- sim_model_from_fits(fits, observed_data, types)
model
```

Each process uses its `coxph()` fit's cumulative baseline hazard as-is, so the
simulated event times follow its shape. The hazard is unknown beyond the last
observed time, so simulated follow-up ends there (`event = "max_cens"`).

```{r}
#| label: sim-model-from-fits-simulate

new_data <- sim_events(model, n = 1000)
head(new_data)

# Event-type distribution should match between observed and simulated data (a
# few simulated individuals reach the end of the observed follow-up, as
# "max_cens"):
event_levels <- levels(new_data$event)
rbind(
  observed = prop.table(table(factor(observed_data$event, event_levels))),
  simulated = prop.table(table(new_data$event))
)
```

## Categorical Baseline Covariates

A categorical baseline covariate must be a factor column in `data`, referenced
directly in the `coxph()` formula (not wrapped in `factor()` there).
`sim_model_from_fits()` regenerates it as its integer level codes, plus one
`sim_derived()` dummy per non-reference level, named to match `coxph()`'s own
coefficient names (`"<variable><level>"`). The model-native replacement for
`sim.from.data()`'s `"(region==2)"`-style expression matching.

```{r}
#| label: sim-model-from-fits-categorical

observed_data$region <- factor(sample(
  c("a", "b", "c"),
  nrow(observed_data),
  replace = TRUE,
  prob = c(0.5, 0.3, 0.2)
))

fits_region <- list(
  censoring = coxph(
    Surv(time, event == "censoring") ~ L0 + region,
    data = observed_data
  ),
  cause1 = coxph(
    Surv(time, event == "cause1") ~ L0 + region,
    data = observed_data
  ),
  cause2 = coxph(Surv(time, event == "cause2") ~ L0, data = observed_data)
)

model_region <- sim_model_from_fits(fits_region, observed_data, types)
model_region
```
