---
title: "Migrating to sim_event_graph()"
vignette: >
  %\VignetteIndexEntry{Migrating to sim_event_graph()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-opts
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup

library(simevent)
library(survival)

compare_graph <- function(wrapper_data, graph_data) {
  wrapper_data <- as.data.frame(wrapper_data)
  graph_data <- as.data.frame(graph_data)
  names(graph_data)[match(c("id", "time", "delta"), names(graph_data))] <-
    c("ID", "Time", "Delta")
  isTRUE(all.equal(
    wrapper_data,
    graph_data[names(wrapper_data)],
    tolerance = 1e-6
  ))
}
```

`compare_graph()` checks that a `sim_event_graph()` reproduction draws the exact
same random numbers, in the same order, as the wrapper it's reproducing. That
only holds below for `simSurvData()`/`simCRdata()`.

`simDisease()`/`simTreatment()` each have a `"transient"` process, so some
individuals finish before others; `sim_event_graph()` draws random numbers only
for whoever's still being simulated, while the wrapper always draws for everyone
(including those already finished) and discards the unused draws.

# Introduction

This introduces the new API with `sim_event_graph()`, which is (probably?) a
better interface.

# The `sim_graph()` Interface

A `sim_graph()` call has three kinds of possible inputs:

- `sim_covariate()`: a baseline covariate generator. Takes `N` and, optionally,
  any other covariate defined earlier in the same call (matched by argument
  name).
- `sim_process()`: a `"censoring"`, `"terminal"`, or `"transient"` (fires at
  most `limit` times, default `Inf`) process, with Weibull `eta`/`nu`
  parameters.
- `sim_effect()`: a `from -> to` edge with a Cox-type `coef`, where `from` is a
  covariate or process name and `to` is a process name.

```{r}
#| label: sim-graph-basic

graph <- sim_graph(
  L0 = sim_covariate(function(N) rbinom(N, 1, 0.4)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1),
  effects = list(sim_effect("L0", "death", coef = 1.5))
)
graph

data <- sim_event_graph(graph, n = 100)
head(data)
```

`sim_event_graph()`'s `intervene` argument replaces `sim.generic()`'s
`alpha.intervention`/`baseline.intervention` pair with a single named list: give
a process name to scale its `eta`, or a covariate name to fix it to a constant.

```{r}
#| label: sim-graph-intervention

data_intervened <- sim_event_graph(
  graph,
  n = 100,
  intervene = list(death = 0.2, L0 = 1)
)
head(data_intervened)
```

# Reproducing the Preset Wrappers

## `simSurvData()`

`simSurvData()`'s default (`beta = NULL`, i.e. no covariate effects) is two
terminal processes, censoring and death, both with `eta = 0.1`, `nu = 1.1`:

```{r}
#| label: simsurvdata-equivalence

set.seed(1405)
wrapper_data <- simSurvData(200)

set.seed(1405)
graph <- sim_graph(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1)
)
graph_data <- sim_event_graph(graph, n = 200)

compare_graph(wrapper_data, graph_data)
```

## `simCRdata()`

`simCRdata()` adds a third terminal process (a second competing cause):

```{r}
#| label: simcrdata-equivalence

set.seed(1405)
beta <- matrix(c(0.5, -1, -0.5, 0.5, 0, 0.5), ncol = 3, nrow = 2)
wrapper_data <- simCRdata(N = 200, beta = beta)

set.seed(1405)
graph <- sim_graph(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  cause1 = sim_process("terminal", eta = 0.1, nu = 1.1),
  cause2 = sim_process("terminal", eta = 0.1, nu = 1.1),
  effects = list(
    sim_effect("L0", "censoring", 0.5),
    sim_effect("A0", "censoring", -1),
    sim_effect("L0", "cause1", -0.5),
    sim_effect("A0", "cause1", 0.5),
    sim_effect("A0", "cause2", 0.5)
  )
)
graph_data <- sim_event_graph(graph, n = 200)

compare_graph(wrapper_data, graph_data)
```

## `simDisease()`

`simDisease()` has a censoring process, a terminal death process, and a
covariate process `L` that can happen at most once and can itself affect death;
a `"transient"` process with `limit = 1`. Since `L` can fire, the simulation
loop can run more than once per individual, so `sim_event_graph()`'s random
draws no longer line up one-to-one with `simDisease()`'s (see the note on
`compare_graph()` above); event-type proportions are compared instead:

```{r}
#| label: simdisease-equivalence

set.seed(1405)
wrapper_data <- simDisease(
  N = 5000,
  cens = 1,
  eta = c(0.1, 0.3, 0.1),
  nu = c(1.1, 1.3, 1.1),
  beta_L0_L = 1,
  beta_A0_L = -1.1,
  beta_L_D = 1,
  beta_L0_D = 0
)

graph <- sim_graph(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.3, nu = 1.3),
  L = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
  effects = list(
    sim_effect("L0", "L", 1),
    sim_effect("A0", "L", -1.1),
    sim_effect("L", "death", 1)
  )
)
graph_data <- sim_event_graph(graph, n = 5000)

rbind(
  simDisease = prop.table(table(wrapper_data$Delta)),
  sim_graph = prop.table(table(graph_data$delta))
)
```

## `simTreatment()`

`simTreatment()` (with `op = 1`, the default) adds a second `"transient"`
process, treatment (`A`), which can itself affect and be affected by the
covariate process `L`. `simTreatment()` doesn't declare its own `A0`, so its
reproduction doesn't need one either -- `sim_graph()` forces no default
covariates:

```{r}
#| label: simtreatment-equivalence

set.seed(1405)
wrapper_data <- simTreatment(
  N = 5000,
  beta_L_A = 1,
  beta_L_D = 1,
  beta_A_D = -1,
  beta_A_L = -0.5,
  beta_L0_A = 1,
  eta = rep(0.1, 4),
  nu = rep(1.1, 4),
  cens = 1,
  op = 1
)

graph <- sim_graph(
  L0 = sim_covariate(function(N) runif(N)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1),
  A = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
  L = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
  effects = list(
    sim_effect("L0", "death", 1),
    sim_effect("L0", "A", 1),
    sim_effect("L0", "L", 1),
    sim_effect("A", "death", -1),
    sim_effect("A", "L", -0.5),
    sim_effect("L", "death", 1),
    sim_effect("L", "A", 1)
  )
)
graph_data <- sim_event_graph(graph, n = 5000)

rbind(
  simTreatment = prop.table(table(wrapper_data$Delta)),
  sim_graph = prop.table(table(graph_data$delta))
)
```

## `simStatinData()`

`simStatinData()` mainly demonstrates non-default baseline covariate generators.
`sim_covariate()` reproduces this directly, with no forced covariate names to
work around:

```{r}
#| label: simstatindata-equivalence

set.seed(1405)
graph <- sim_graph(
  L0 = sim_covariate(function(N) rbinom(N, 1, 0.4)),
  A0 = sim_covariate(function(N, L0) pmin(rexp(N, 0.3) + 70, 100)),
  censoring = sim_process("censoring", eta = 0.025, nu = 1.1),
  death = sim_process("terminal", eta = 0.025, nu = 1.1)
)
graph_data <- sim_event_graph(graph, n = 200)
summary(graph_data$A0)
```

# Estimating Intervention Effects

`alphaSim()` and `intEffectAlpha()` simulate one of the preset settings with the
shape parameter `eta` of one process multiplied by `alpha`, and summarise the
result as the proportion of individuals experiencing death and the intervened
process by time $\tau$ (or the years lost to them before $\tau$). With the graph
API this splits into two general steps, which work for any process in any graph:

1. Simulate under the intervention with `sim_event_graph()`'s `intervene`
   argument: `intervene = list(<process> = alpha)` multiplies that process's
   `eta` by `alpha`.
2. Summarise with `event_risk()`: the risk $P(T \le \tau)$ of each process's
   first event, or (`type = "time_lost"`) the expected time lost to it before
   $\tau$, $E[\tau - \min(T, \tau)]$.

Take `alphaSim()`'s `"Disease"` setting with `alpha = 0.5`, which halves the
hazard of the disease process `L`:

```{r}
#| label: alphasim-old

set.seed(1405)
alphaSim(
  N = 5000,
  eta = rep(0.1, 3),
  nu = rep(1.1, 3),
  alpha = 0.5,
  tau = 5,
  setting = "Disease"
)
```

The same model as a graph:

```{r}
#| label: intervention-graph

disease_graph <- sim_graph(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1),
  L = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
  effects = list(
    sim_effect("L0", "death", 1),
    sim_effect("L0", "L", 1),
    sim_effect("L", "death", 1)
  )
)
```

`alphaSim()` switches censoring off by default (`cens = 0`). Simulating both
with and without the intervention gives the effect directly:

```{r}
#| label: intervention-risk

set.seed(1405)
observed <- sim_event_graph(disease_graph, n = 5000, cens = 0)
intervened <- sim_event_graph(
  disease_graph,
  n = 5000,
  cens = 0,
  intervene = list(L = 0.5)
)

risk_observed <- event_risk(observed, disease_graph, c("death", "L"), tau = 5)
risk_intervened <- event_risk(
  intervened,
  disease_graph,
  c("death", "L"),
  tau = 5
)
cbind(
  risk_observed[, "process"],
  observed = risk_observed$risk,
  intervened = risk_intervened$risk
)
```

Halving the disease hazard lowers the risk of disease, and, since disease raises
the death hazard (`sim_effect("L", "death", 1)`), the risk of death too.

`alphaSim()`/`intEffectAlpha()`'s `years_lost = TRUE` corresponds to
`type = "time_lost"`, and their `a0` argument (summarise only individuals with
`A0 == a0`) to `by = "A0"`, which summarises every group at once:

```{r}
#| label: intervention-time-lost

event_risk(
  intervened,
  disease_graph,
  c("death", "L"),
  tau = 5,
  type = "time_lost",
  by = "A0"
)
```

`by` conditions on the observed value of `A0`. For a `do()`-style contrast,
where everyone's `A0` is set to a value, fix it with `intervene` instead, e.g.
`intervene = list(L = 0.5, A0 = 1)`.

Finally, `intEffectAlpha()`'s `plot = TRUE` is `plot_event_data()` on the
intervened data:

```{r}
#| label: intervention-plot

plot_event_data(
  intervened[intervened$id <= 100, ],
  title = "Under Intervention"
)
```

# Building a `sim_graph()` from Fitted Cox Models

`sim_graph_from_fits()` is the graph-native replacement for `sim.from.data()`:
instead of user-specified effects, it builds a `sim_graph()` from a set of
fitted `coxph()` models (one per process) and the data they were fit to, so
simulated data mimics an observed dataset's distribution.

Unlike `sim.from.data()`, which requires building a `sim.parameters` list,
including a `baseline.summary` computed by hand from `mean()`/
`sd()`/`min()`/`max()`, `sim_graph_from_fits()` reads baseline covariate
distributions straight from the data itself.

```{r}
#| label: sim-graph-from-fits-setup

# Some "observed" data, from a 3-cause competing-risks sim_graph():
set.seed(1405)
observed_graph <- sim_graph(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  cause1 = sim_process("terminal", eta = 0.1, nu = 1.1),
  cause2 = sim_process("terminal", eta = 0.1, nu = 1.1),
  effects = list(
    sim_effect("L0", "censoring", 0.5),
    sim_effect("A0", "censoring", -1),
    sim_effect("L0", "cause1", -0.5),
    sim_effect("A0", "cause1", 0.5),
    sim_effect("A0", "cause2", 0.5)
  )
)
observed_data <- sim_event_graph(observed_graph, n = 1000)

# Refit each process from that "observed" data, then rebuild a sim_graph()
# from the fits, as if observed_data came from an outside source and
# observed_graph were unknown:
fits <- list(
  censoring = coxph(Surv(time, delta == 0) ~ L0 + A0, data = observed_data),
  cause1 = coxph(Surv(time, delta == 1) ~ L0 + A0, data = observed_data),
  cause2 = coxph(Surv(time, delta == 2) ~ L0 + A0, data = observed_data)
)
types <- c(censoring = "censoring", cause1 = "terminal", cause2 = "terminal")

graph <- sim_graph_from_fits(fits, observed_data, types)
graph
```

Each process's Weibull `eta`/`nu` is approximated from its `coxph()` fit's
baseline cumulative hazard, by regressing log(cumulative hazard) on log(time) (a
Weibull hazard is linear on that scale).

```{r}
#| label: sim-graph-from-fits-simulate

new_data <- sim_event_graph(graph, n = 1000)
head(new_data)

# Event-type distribution should be comparable between the observed and
# newly simulated data:
rbind(
  observed = prop.table(table(observed_data$delta)),
  simulated = prop.table(table(new_data$delta))
)
```

## Categorical Baseline Covariates

A categorical baseline covariate must be a factor column in `data`, referenced
directly in the `coxph()` formula (not wrapped in `factor()` there).
`sim_graph_from_fits()` regenerates it as a categorical draw from its observed
level proportions, plus one `sim_derived()` dummy per non-reference level, named
to match `coxph()`'s own coefficient names (`"<variable><level>"`). The
graph-native replacement for `sim.from.data()`'s `"(region==2)"`-style
expression matching.

```{r}
#| label: sim-graph-from-fits-categorical

observed_data$region <- factor(sample(
  c("a", "b", "c"),
  nrow(observed_data),
  replace = TRUE,
  prob = c(0.5, 0.3, 0.2)
))

fits_region <- list(
  censoring = coxph(
    Surv(time, delta == 0) ~ L0 + region,
    data = observed_data
  ),
  cause1 = coxph(Surv(time, delta == 1) ~ L0 + region, data = observed_data),
  cause2 = coxph(Surv(time, delta == 2) ~ L0, data = observed_data)
)

graph_region <- sim_graph_from_fits(fits_region, observed_data, types)
graph_region
```
