---
title: "Migrating to sim.generic()"
vignette: >
  %\VignetteIndexEntry{Migrating to sim.generic()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-opts
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup

library(simevent)
library(survival)

# Compares a preset wrapper's output to sim.generic()'s, up to column order
# and floating-point root-finding precision (the two use slightly different
# default `lower`/`upper` root-finding bounds).
compare_wrapper <- function(wrapper_data, generic_data) {
  wrapper_data <- as.data.frame(wrapper_data)
  generic_data <- as.data.frame(setNames(
    generic_data,
    names(wrapper_data)
  ))[names(wrapper_data)]
  isTRUE(all.equal(wrapper_data, generic_data, tolerance = 1e-6))
}
```

# Introduction

`vignette("simevent")` introduces `simEventData()` and a set of wrapper
functions built on top of it for common settings: `simSurvData()`,
`simCRdata()`, `simDisease()`, `simTreatment()`, `simDropIn()`, and
`simStatinData()`. Each of these wrappers just builds a fixed `beta`/`eta`/
`nu`/`at_risk` specification and calls `simEventData()`.

`sim.generic()` is a general-purpose alternative: instead of a raw `beta` matrix
indexed by column position, you describe a named list of baseline covariates, a
named list of processes (with a `type` and Weibull `eta`/`nu` parameters), and a
list of `c(from, to, coefficient)` effects between them.

# The `sim.generic()` Interface

A `sim.generic()` call has three parts:

- `baseline`: a named list of baseline covariate generators. Any name other than
  `L0`/`A0` becomes an extra baseline covariate in the output.
- `processes`: a named list of process specifications, each a list with a `type`
  (`"censoring"`, `"terminal"`, `"one.jump"`, or anything else for a recurrent
  process) and Weibull `eta`/`nu` parameters.
- `effects`: a list of `c(from, to, coefficient)` triples, where `from` is a
  baseline covariate or process name and `to` is a process name.

```{r}
#| label: sim-generic-basic

baseline <- list(L0 = function(N) rbinom(N, 1, 0.4))
processes <- list(
  censoring = list(type = "censoring", eta = 0.1, nu = 1.1),
  death = list(type = "terminal", eta = 0.1, nu = 1.1)
)
effects <- list(c("L0", "death", 1.5))

data <- sim.generic(baseline, processes, effects, n = 100)
head(data)
```

`sim.generic()` also supports `alpha.intervention` (a named list of
multiplicative changes to a process's `eta`) and `baseline.intervention` (a
named list fixing a baseline covariate to a constant), for simulating under an
intervention:

```{r}
#| label: sim-generic-intervention

data_intervened <- sim.generic(
  baseline,
  processes,
  effects,
  alpha.intervention = list(death = 0.2),
  baseline.intervention = list(L0 = 1),
  n = 100
)
head(data_intervened)
```

# Reproducing the Preset Wrappers

## `simSurvData()`

`simSurvData()`'s default (`beta = NULL`, i.e. no covariate effects) is two
terminal processes, censoring and death, both with `eta = 0.1`, `nu = 1.1`:

```{r}
#| label: simsurvdata-equivalence

set.seed(11)
wrapper_data <- simSurvData(200)

set.seed(11)
generic_data <- sim.generic(
  baseline = list(
    L0 = function(N) runif(N),
    A0 = function(N, L0) rbinom(N, 1, 0.5)
  ),
  processes = list(
    censoring = list(type = "censoring", eta = 0.1, nu = 1.1),
    death = list(type = "terminal", eta = 0.1, nu = 1.1)
  ),
  effects = list(),
  n = 200
)

compare_wrapper(wrapper_data, generic_data)
```

## `simCRdata()`

`simCRdata()` adds a third terminal process (a second competing cause):

```{r}
#| label: simcrdata-equivalence

set.seed(22)
beta <- matrix(c(0.5, -1, -0.5, 0.5, 0, 0.5), ncol = 3, nrow = 2)
wrapper_data <- simCRdata(N = 200, beta = beta)

set.seed(22)
generic_data <- sim.generic(
  baseline = list(
    L0 = function(N) runif(N),
    A0 = function(N, L0) rbinom(N, 1, 0.5)
  ),
  processes = list(
    censoring = list(type = "censoring", eta = 0.1, nu = 1.1),
    cause1 = list(type = "terminal", eta = 0.1, nu = 1.1),
    cause2 = list(type = "terminal", eta = 0.1, nu = 1.1)
  ),
  effects = list(
    c("L0", "censoring", 0.5),
    c("A0", "censoring", -1),
    c("L0", "cause1", -0.5),
    c("A0", "cause1", 0.5),
    c("A0", "cause2", 0.5)
  ),
  n = 200
)

compare_wrapper(wrapper_data, generic_data)
```

Each `c(from, to, coefficient)` triple is one non-zero cell of `beta`: cell
`beta[row, col]` becomes
`c(rownames(beta)[row], colnames(beta)[col], beta[row, col])`.

## `simDisease()`

`simDisease()` has a censoring process, a terminal death process, and a
`one.jump` covariate process `L` (an event that can happen at most once and can
itself affect death):

```{r}
#| label: simdisease-equivalence

set.seed(6258)
wrapper_data <- simDisease(
  N = 200,
  cens = 1,
  eta = c(0.1, 0.3, 0.1),
  nu = c(1.1, 1.3, 1.1),
  beta_L0_L = 1,
  beta_A0_L = -1.1,
  beta_L_D = 1,
  beta_L0_D = 0
)

set.seed(6258)
generic_data <- sim.generic(
  baseline = list(
    L0 = function(N) runif(N),
    A0 = function(N, L0) rbinom(N, 1, 0.5)
  ),
  processes = list(
    censoring = list(type = "censoring", eta = 0.1, nu = 1.1),
    death = list(type = "terminal", eta = 0.3, nu = 1.3),
    L = list(type = "one.jump", eta = 0.1, nu = 1.1)
  ),
  effects = list(
    c("L0", "L", 1),
    c("A0", "L", -1.1),
    c("L", "death", 1)
  ),
  n = 200
)

compare_wrapper(wrapper_data, generic_data)
```

A `one.jump` process is only at risk while its own event count is still 0.

## `simTreatment()`

`simTreatment()` (with `op = 1`, the default) adds a second `one.jump` process,
treatment (`A`), which can itself affect and be affected by the covariate
process `L`:

```{r}
#| label: simtreatment-equivalence

set.seed(6258)
wrapper_data <- simTreatment(
  N = 200,
  beta_L_A = 1,
  beta_L_D = 1,
  beta_A_D = -1,
  beta_A_L = -0.5,
  beta_L0_A = 1,
  eta = rep(0.1, 4),
  nu = rep(1.1, 4),
  cens = 1,
  op = 1
)

set.seed(6258)
generic_data <- sim.generic(
  baseline = list(L0 = function(N) runif(N)),
  processes = list(
    censoring = list(type = "censoring", eta = 0.1, nu = 1.1),
    death = list(type = "terminal", eta = 0.1, nu = 1.1),
    A = list(type = "one.jump", eta = 0.1, nu = 1.1),
    L = list(type = "one.jump", eta = 0.1, nu = 1.1)
  ),
  effects = list(
    c("L0", "death", 1),
    c("L0", "A", 1),
    c("L0", "L", 1),
    c("A", "death", -1),
    c("A", "L", -0.5),
    c("L", "death", 1),
    c("L", "A", 1)
  ),
  n = 200
)

compare_wrapper(wrapper_data, generic_data)
```

## `simStatinData()`

`simStatinData()` mainly demonstrates non-default baseline covariate generators.
The `baseline` argument reproduces this directly:

```{r}
#| label: simstatindata-equivalence

set.seed(4)
generic_data <- sim.generic(
  baseline = list(
    L0 = function(N) rbinom(N, 1, 0.4),
    A0 = function(N, L0) pmin(rexp(N, 0.3) + 70, 100)
  ),
  processes = list(
    censoring = list(type = "censoring", eta = 0.025, nu = 1.1),
    death = list(type = "terminal", eta = 0.025, nu = 1.1)
  ),
  effects = list(),
  n = 200
)
summary(generic_data$A0)
```

# Simulating New Data from Fitted Parameters with `sim.from.data()`

`sim.from.data()` is like `sim.generic()`, but instead of user-specified
effects, it takes `sim.parameters`: Weibull and Cox parameters previously
*estimated* from observed data.

`sim.parameters` is a named list with one entry per process, plus a
`baseline.summary` entry and a `model.structure` entry. Building it requires
fitting a `coxph()` model per process and extracting a Weibull approximation
from its baseline cumulative hazard; a small helper does the latter:

```{r}
#| label: sim-from-data-setup

set.seed(373)
beta <- matrix(c(0.5, -1, -0.5, 0.5, 0, 0.5), ncol = 3, nrow = 2)
observed_data <- simCRdata(N = 1000, beta = beta)

# Approximate a Cox model's baseline cumulative hazard with a Weibull curve,
# by regressing log(cumulative hazard) on log(time).
weibull_params <- function(fit) {
  bh <- basehaz(fit, centered = FALSE)
  bh <- bh[bh$hazard > 0, ]
  fit_weibull <- lm(log(hazard) ~ log(time), data = bh)
  c(eta = unname(exp(coef(fit_weibull)[1])), nu = unname(coef(fit_weibull)[2]))
}

cox_fits <- list(
  censoring = coxph(Surv(Time, Delta == 0) ~ L0 + A0, data = observed_data),
  cause1 = coxph(Surv(Time, Delta == 1) ~ L0 + A0, data = observed_data),
  cause2 = coxph(Surv(Time, Delta == 2) ~ L0 + A0, data = observed_data)
)

sim.parameters <- lapply(cox_fits, function(fit) {
  list(weibull.parameters = weibull_params(fit), cox.parameters = coef(fit))
})

sim.parameters$baseline.summary <- list(
  L0 = list(
    type = "numeric",
    mean = mean(observed_data$L0),
    sd = sd(observed_data$L0),
    min = min(observed_data$L0),
    max = max(observed_data$L0)
  ),
  A0 = list(
    type = "numeric",
    mean = mean(observed_data$A0),
    sd = sd(observed_data$A0),
    min = min(observed_data$A0),
    max = max(observed_data$A0)
  )
)

sim.parameters$model.structure <- list(
  process.names = c("censoring", "cause1", "cause2"),
  process.deltas = c(0, 1, 2),
  process.types = c("terminal", "terminal", "terminal"),
  cens.process.id = 1
)
```

`baseline.summary`'s `type = "numeric"` entries are treated as Bernoulli if
`min`/`max` are exactly `0`/`1` (as `L0` is here), and Normal otherwise (as `A0`
is here); a `type` other than `"numeric"` is treated as categorical, drawing
from `proportions` (a named list/table of category probabilities) instead of
`mean`/`sd`.

```{r}
#| label: sim-from-data-simulate

new_data <- sim.from.data(n = 1000, sim.parameters = sim.parameters)
head(new_data)

# Event-type distribution should be comparable between the observed and
# newly simulated data:
rbind(
  observed = prop.table(table(observed_data$Delta)),
  simulated = prop.table(table(new_data$delta))
)
```
