---
title: "Migrating to sim_event_graph()"
vignette: >
  %\VignetteIndexEntry{Migrating to sim_event_graph()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-opts
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup

library(simevent)
library(survival)

compare_graph <- function(wrapper_data, graph_data) {
  wrapper_data <- as.data.frame(wrapper_data)
  graph_data <- as.data.frame(graph_data)
  names(graph_data)[match(c("id", "time", "delta"), names(graph_data))] <-
    c("ID", "Time", "Delta")
  isTRUE(all.equal(
    wrapper_data,
    graph_data[names(wrapper_data)],
    tolerance = 1e-6
  ))
}
```

# Introduction

This introduces the new API with `sim_event_graph()`, which is (probably?) a
better interface.

# The `sim_graph()` Interface

A `sim_graph()` call has three kinds of possible inputs:

- `sim_covariate()`: a baseline covariate generator. Takes `N` and, optionally,
  any other covariate defined earlier in the same call (matched by argument
  name).
- `sim_process()`: a `"censoring"`, `"terminal"`, or `"transient"` (fires at
  most `limit` times, default `Inf`) process, with Weibull `eta`/`nu`
  parameters.
- `sim_effect()`: a `from -> to` edge with a Cox-type `coef`, where `from` is a
  covariate or process name and `to` is a process name.

```{r}
#| label: sim-graph-basic

graph <- sim_graph(
  L0 = sim_covariate(function(N) rbinom(N, 1, 0.4)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1),
  effects = list(sim_effect("L0", "death", coef = 1.5))
)
graph

data <- sim_event_graph(graph, n = 100)
head(data)
```

`sim_event_graph()`'s `intervene` argument replaces `sim.generic()`'s
`alpha.intervention`/`baseline.intervention` pair with a single named list: give
a process name to scale its `eta`, or a covariate name to fix it to a constant.

```{r}
#| label: sim-graph-intervention

data_intervened <- sim_event_graph(
  graph,
  n = 100,
  intervene = list(death = 0.2, L0 = 1)
)
head(data_intervened)
```

# Reproducing the Preset Wrappers

## `simSurvData()`

`simSurvData()`'s default (`beta = NULL`, i.e. no covariate effects) is two
terminal processes, censoring and death, both with `eta = 0.1`, `nu = 1.1`:

```{r}
#| label: simsurvdata-equivalence

set.seed(1405)
wrapper_data <- simSurvData(200)

set.seed(1405)
graph <- sim_graph(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1)
)
graph_data <- sim_event_graph(graph, n = 200)

compare_graph(wrapper_data, graph_data)
```

## `simCRdata()`

`simCRdata()` adds a third terminal process (a second competing cause):

```{r}
#| label: simcrdata-equivalence

set.seed(1405)
beta <- matrix(c(0.5, -1, -0.5, 0.5, 0, 0.5), ncol = 3, nrow = 2)
wrapper_data <- simCRdata(N = 200, beta = beta)

set.seed(1405)
graph <- sim_graph(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  cause1 = sim_process("terminal", eta = 0.1, nu = 1.1),
  cause2 = sim_process("terminal", eta = 0.1, nu = 1.1),
  effects = list(
    sim_effect("L0", "censoring", 0.5),
    sim_effect("A0", "censoring", -1),
    sim_effect("L0", "cause1", -0.5),
    sim_effect("A0", "cause1", 0.5),
    sim_effect("A0", "cause2", 0.5)
  )
)
graph_data <- sim_event_graph(graph, n = 200)

compare_graph(wrapper_data, graph_data)
```

## `simDisease()`

`simDisease()` has a censoring process, a terminal death process, and a
covariate process `L` that can happen at most once and can itself affect death;
a `"transient"` process with `limit = 1`:

```{r}
#| label: simdisease-equivalence

set.seed(1405)
wrapper_data <- simDisease(
  N = 200,
  cens = 1,
  eta = c(0.1, 0.3, 0.1),
  nu = c(1.1, 1.3, 1.1),
  beta_L0_L = 1,
  beta_A0_L = -1.1,
  beta_L_D = 1,
  beta_L0_D = 0
)

set.seed(1405)
graph <- sim_graph(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.3, nu = 1.3),
  L = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
  effects = list(
    sim_effect("L0", "L", 1),
    sim_effect("A0", "L", -1.1),
    sim_effect("L", "death", 1)
  )
)
graph_data <- sim_event_graph(graph, n = 200)

compare_graph(wrapper_data, graph_data)
```

## `simTreatment()`

`simTreatment()` (with `op = 1`, the default) adds a second `"transient"`
process, treatment (`A`), which can itself affect and be affected by the
covariate process `L`:

```{r}
#| label: simtreatment-equivalence

set.seed(1405)
wrapper_data <- simTreatment(
  N = 200,
  beta_L_A = 1,
  beta_L_D = 1,
  beta_A_D = -1,
  beta_A_L = -0.5,
  beta_L0_A = 1,
  eta = rep(0.1, 4),
  nu = rep(1.1, 4),
  cens = 1,
  op = 1
)

set.seed(1405)
graph <- sim_graph(
  L0 = sim_covariate(function(N) runif(N)),
  # simTreatment() doesn't declare its own A0, so it silently relies on
  # simEventData()'s forced default A0 draw (Bernoulli(0.5)) purely to
  # consume the same random numbers, then discards it.
  # sim_graph() forces nothing, so reproducing simTreatment()'s
  # random numbers bit-for-bit means declaring that same unused A0
  # explicitly.
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  death = sim_process("terminal", eta = 0.1, nu = 1.1),
  A = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
  L = sim_process("transient", eta = 0.1, nu = 1.1, limit = 1),
  effects = list(
    sim_effect("L0", "death", 1),
    sim_effect("L0", "A", 1),
    sim_effect("L0", "L", 1),
    sim_effect("A", "death", -1),
    sim_effect("A", "L", -0.5),
    sim_effect("L", "death", 1),
    sim_effect("L", "A", 1)
  )
)
graph_data <- sim_event_graph(graph, n = 200)

compare_graph(wrapper_data, graph_data)
```

## `simStatinData()`

`simStatinData()` mainly demonstrates non-default baseline covariate generators.
`sim_covariate()` reproduces this directly, with no forced covariate names to
work around:

```{r}
#| label: simstatindata-equivalence

set.seed(1405)
graph <- sim_graph(
  L0 = sim_covariate(function(N) rbinom(N, 1, 0.4)),
  A0 = sim_covariate(function(N, L0) pmin(rexp(N, 0.3) + 70, 100)),
  censoring = sim_process("censoring", eta = 0.025, nu = 1.1),
  death = sim_process("terminal", eta = 0.025, nu = 1.1)
)
graph_data <- sim_event_graph(graph, n = 200)
summary(graph_data$A0)
```

# Building a `sim_graph()` from Fitted Cox Models

`sim_graph_from_fits()` is the graph-native replacement for `sim.from.data()`:
instead of user-specified effects, it builds a `sim_graph()` from a set of
fitted `coxph()` models (one per process) and the data they were fit to, so
simulated data mimics an observed dataset's distribution.

Unlike `sim.from.data()`, which requires building a `sim.parameters` list,
including a `baseline.summary` computed by hand from `mean()`/
`sd()`/`min()`/`max()`, `sim_graph_from_fits()` reads baseline covariate
distributions straight from the data itself.

```{r}
#| label: sim-graph-from-fits-setup

# Some "observed" data, from a 3-cause competing-risks sim_graph():
set.seed(1405)
observed_graph <- sim_graph(
  L0 = sim_covariate(function(N) runif(N)),
  A0 = sim_covariate(function(N, L0) rbinom(N, 1, 0.5)),
  censoring = sim_process("censoring", eta = 0.1, nu = 1.1),
  cause1 = sim_process("terminal", eta = 0.1, nu = 1.1),
  cause2 = sim_process("terminal", eta = 0.1, nu = 1.1),
  effects = list(
    sim_effect("L0", "censoring", 0.5),
    sim_effect("A0", "censoring", -1),
    sim_effect("L0", "cause1", -0.5),
    sim_effect("A0", "cause1", 0.5),
    sim_effect("A0", "cause2", 0.5)
  )
)
observed_data <- sim_event_graph(observed_graph, n = 1000)

# Refit each process from that "observed" data, then rebuild a sim_graph()
# from the fits, as if observed_data came from an outside source and
# observed_graph were unknown:
fits <- list(
  censoring = coxph(Surv(time, delta == 0) ~ L0 + A0, data = observed_data),
  cause1 = coxph(Surv(time, delta == 1) ~ L0 + A0, data = observed_data),
  cause2 = coxph(Surv(time, delta == 2) ~ L0 + A0, data = observed_data)
)
types <- c(censoring = "censoring", cause1 = "terminal", cause2 = "terminal")

graph <- sim_graph_from_fits(fits, observed_data, types)
graph
```

Each process's Weibull `eta`/`nu` is approximated from its `coxph()` fit's
baseline cumulative hazard, the same way `vignette("sim-generic")`'s
`weibull_params()` helper does it by hand.

```{r}
#| label: sim-graph-from-fits-simulate

new_data <- sim_event_graph(graph, n = 1000)
head(new_data)

# Event-type distribution should be comparable between the observed and
# newly simulated data:
rbind(
  observed = prop.table(table(observed_data$delta)),
  simulated = prop.table(table(new_data$delta))
)
```

## Categorical Baseline Covariates

A categorical baseline covariate must be a factor column in `data`, referenced
directly in the `coxph()` formula (not wrapped in `factor()` there).
`sim_graph_from_fits()` regenerates it as a categorical draw from its observed
level proportions, plus one `sim_derived()` dummy per non-reference level, named
to match `coxph()`'s own coefficient names (`"<variable><level>"`). The
graph-native replacement for `sim.from.data()`'s `"(region==2)"`-style
expression matching.

```{r}
#| label: sim-graph-from-fits-categorical

observed_data$region <- factor(sample(
  c("a", "b", "c"),
  nrow(observed_data),
  replace = TRUE,
  prob = c(0.5, 0.3, 0.2)
))

fits_region <- list(
  censoring = coxph(
    Surv(time, delta == 0) ~ L0 + region,
    data = observed_data
  ),
  cause1 = coxph(Surv(time, delta == 1) ~ L0 + region, data = observed_data),
  cause2 = coxph(Surv(time, delta == 2) ~ L0, data = observed_data)
)

graph_region <- sim_graph_from_fits(fits_region, observed_data, types)
graph_region
```
