---
title: "Getting Started with simevent"
vignette: >
  %\VignetteIndexEntry{Getting Started with simevent}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-opts
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup

library(simevent)
```

# Introduction

The goal of simevent is to provide functions for the generation and analysis
of complex continuous time health care data. The simulation functions are quite
general and many various settings can be specified. In general data is simulated
for $N$ individuals, and the simulated data can include variables such as
treatment decisions, disease progression, and health factors. At the core of
the package is the general-purpose function `simEventData()`, on top of which the
package builds wrapper functions for common settings (e.g. survival data, competing risks, disease progression), functions for simulating new data from
models already fitted to observed data, and functions for performing and
evaluating interventions.

This document first explains the general simulation setting and procedure —
this section is fairly mathematical and aims to give an in-depth account of how
the simulation works — before walking through the package's functions and how
to use them.

# The General Simulation Setting

The simulation approach is similar to [Eimermacher et al., 2015](https://doi.org/10.1186/s12874-015-0005-2), and is based on a counting process framework.

## The Mathematics

We consider a scenario where $n$ individuals are followed over a time interval
$\mathcal{T} = [0, \rho)$, where $\rho \in (0,\infty]$. Each individual can
experience $J \geq 1$ events. Let $\mathcal{X} = \lbrace 0, 1, \ldots , J \rbrace$ represent the various types of events. We often let $x = 0$ denote
censoring. Events are not necessarily exclusive and can affect one another. We
let $N^x$ represent the counting process associated to the event $x \in
\mathcal{X}$. We collect all the counting processes in a $J + 1$ dimensional
vector $N$. The filtration $(\mathcal{F}_t)_{t \in \mathcal{T}}$ is
generated by the variables up to time $t$, and represents the information
available at this time point. The filtration $\mathcal{F}_{t-}$ represents
the information available just before the time $t$.

We assume the intensity of the counting processes takes the form,
$$
  \lambda^x(t\, \vert \, \mathcal{F}_{t-})
  = R^x(t\, \vert \, \mathcal{F}_{t-}) \eta^x \nu^x t^{\nu^x-1} \phi^x(t \, \vert \, \mathcal{F}_{t-})
$$
where $\eta^x > 0$ is a scale parameter and $\nu^x > 0$ is a shape
parameter. This is the intensity corresponding to a Weibull distributed random
variable. When $\nu^x > 1$, the intensity increases over time, and when
$\nu^x < 1$, the intensity decreases. The function $R^x(t \, \vert \, \mathcal{F}*{t-})$ is an ($x$-specific) at-risk indicator that determines
when an individual is at risk for event $x$. The function
$\phi^x(t \, \vert \, \mathcal{F}_{t-})$ models how the intensity depends on past information. We
consider functions of the form:
$$
  \phi^x(t \, \vert \, \mathcal{F}_{t-})
  = \exp (L^{\top}\beta_1 + N(t-)^{\top} \beta_2)
$$
Where $\beta_1 \in \mathbb{R}^d$, $\beta_2 \in \mathbb{R}^{J+1}$ and
$L \in \mathbb{R}^d$ is a vector of baseline covariates.

The simulations are made by deriving the distribution of the waiting times, the
times in between events, and these follow directly from the specification of the
intensities of the originnal counting processes.

## The `simEventData()` Function

`simEventData()` is the general-purpose function implementing the simulation
scheme above. It is quite flexible and therefore takes many arguments; see
`?simEventData()` for the full list. The number of simulated event types
corresponds to the length of the vectors `eta`, `nu`, or the number of columns
in `beta`.

Below, we create a 9x5 `beta` matrix for 5 event processes and 4 baseline
covariates (`L0`, `A0`, and the two additional covariates added below):

```{r}
#| label: beta-matrix

set.seed(736)
beta <- matrix(rnorm(9 * 5), ncol = 5, nrow = 9)
```

You can add baseline covariates by supplying generator functions:

```{r}
#| label: add-cov

func1 <- function(N) rbinom(N, 1, 0.2)
func2 <- function(N) rnorm(N)
add_cov <- list("Z1" = func1, "Z2" = func2)
```

Define an `at_risk()` function to specify risk windows for recurrent events,
based on how many times each event has already occurred:

```{r}
#| label: at-risk

at_risk <- function(events) {
  return(c(
    1,
    1, # Always at risk for event 0 and 1
    as.numeric(events[3] < 2), # Event 2 can occur twice
    as.numeric(events[4] < 1), # Event 3 can occur once
    as.numeric(events[5] < 2) # Event 4 can occur twice
  ))
}
```

Optionally, an `at_risk_cov()` function specifies risk windows as a function of
baseline covariates instead:

```{r}
#| label: at-risk-cov

at_risk_cov <- function(covariates) {
  return(c(
    1,
    1, # The covariates do not change the risk of event 0 and 1
    as.numeric(covariates[1] < 0.5), # Only at risk of event 2 if L0 < 0.5
    as.numeric(covariates[2] == 1), # Only at risk of event 3 if A0 = 1
    1 # The covariates do not change the risk of event 4
  ))
}
```

Simulate data for 5000 individuals:

```{r}
#| label: simulate-data

set.seed(973)
data <- simEventData(
  N = 5000,
  beta = beta,
  add_cov = add_cov,
  at_risk = at_risk,
  at_risk_cov = at_risk_cov
)
head(data)
```

### Formatting Data for Cox Regression

Transform the data into *tstart-tstop* format with `IntFormatData()` (specify
which columns hold the counting-process data), then fit Cox models with the survival package:

```{r}
#| label: cox-fits

library(survival)
library(data.table)
library(ggplot2)

data_int <- IntFormatData(data, N_cols = 8:12)

# Process 0
survfit0 <- coxph(
  Surv(tstart, tstop, Delta == 0) ~ L0 + A0 + Z1 + Z2 + N2 + N3 + N4,
  data = data_int
)

# Process 4
survfit4 <- coxph(
  Surv(tstart, tstop, Delta == 4) ~ L0 + Z1 + Z2 + N2 + N3 + N4,
  data = data_int[(N4 < 2) & (A0 == 1)]
)
```

Visualize the estimated coefficients alongside the true values used to
simulate the data:

```{r}
#| label: plot-cis

CIs <- cbind(
  "Par" = c(
    paste0(rownames(confint(survfit0)), "_N0"),
    paste0(rownames(confint(survfit4)), "_N4")
  ),
  rbind(confint(survfit0), confint(survfit4))
)

rownames(CIs) <- NULL
colnames(CIs) <- c("Par", "Lower", "Upper")
CIs <- data.table(CIs)
CIs[, True_val := c(beta[-c(5, 6), 1], beta[-c(2, 5, 6), 5])]

CIs$Lower <- as.numeric(CIs$Lower)
CIs$Upper <- as.numeric(CIs$Upper)

ggplot(data = CIs) +
  geom_point(aes(y = Par, x = Lower)) +
  geom_point(aes(y = Par, x = Upper)) +
  geom_point(aes(y = Par, x = True_val, col = "Estimate")) +
  scale_color_manual(values = c("Estimate" = "red")) +
  geom_segment(aes(y = Par, yend = Par, x = Lower, xend = Upper)) +
  xlab("Estimates")
```

### Visualizing Event Histories

`plotEventData()` plots individual event histories over time, e.g. for the first 100 individuals:

```{r}
#| label: plot-event-histories

plotEventData(data[1:100, ])
```

### Overriding Entries of the Beta Matrix

The `override_beta` argument lets you set specific effects without rebuilding
the whole `beta` matrix. For example, zero out the effect of `L0` on
processes `N0` and `N1`:

```{r}
#| label: override-beta-zero

data <- simEventData(
  N = 1000,
  beta = beta,
  add_cov = add_cov,
  at_risk = at_risk,
  override_beta = list("L0" = c("N0" = 0, "N1" = 0))
)
```

You can also specify non-linear effects, e.g. increase the effect of `N2` on
`N1` by 2 once `N2 > 1`:

```{r}
#| label: override-beta-nonlinear

data <- simEventData(
  N = 1000,
  beta = beta,
  add_cov = add_cov,
  at_risk = at_risk,
  override_beta = list("N2 > 1" = c("N1" = 2))
)
```

Interaction effects are specified similarly, e.g. an interaction effect of 1
between `L0` and `N3` on `N1`:

```{r}
#| label: override-beta-interaction

data <- simEventData(
  N = 3 * 10^4,
  beta = beta,
  add_cov = add_cov,
  at_risk = at_risk,
  override_beta = list("L0 * N3" = c("N1" = 1))
)
```

And effects that depend on time, e.g. an additional effect of 0.4 on `N2`
once `T1 > 1`:

```{r}
#| label: override-beta-time

data <- simEventData(
  N = 10000,
  beta = beta,
  add_cov = add_cov,
  at_risk = at_risk,
  override_beta = list("T1 > 1" = c("N2" = 1))
)
```

# Wrapper Functions for Common Settings

`simEventData()` is fully general, but for common settings the package provides
ready-made wrappers built on top of it.

## Survival Data with `simSurvData()`

`simSurvData()` simulates survival data for individuals at risk of both
censoring (0) and an event (1). By default there are no covariate effects:

```{r}
#| label: simsurv-default

data <- simSurvData(100)
plotEventData(data, title = "Survival Data")
```

You can specify how the baseline covariates `L0` and `A0` affect the
censoring and event hazards through the `beta` matrix, and set the Weibull
shape ($\eta$) and scale ($\nu$) parameters:

```{r}
#| label: simsurv-custom

# Effects of L0 and A0 on the censoring hazard (1st column) and death hazard (2nd column)
beta <- cbind("Censoring" = c(0, 0), "Death" = c(1, -1))
eta <- c(0.2, 0.2)
nu <- c(1.05, 1.05)

data <- simSurvData(100, beta = beta, eta = eta, nu = nu)
plotEventData(data, title = "Survival Data")
```

## Competing Risk Data with `simCRdata()`

`simCRdata()` simulates data from a competing risk setting with two competing
causes, using the same kind of parameters as `simSurvData()`:

```{r}
#| label: simcrdata

data <- simCRdata(100)
plotEventData(data, title = "Competing Risk Data")
```

## Health Care Data with `simDisease()`

`simDisease()` simulates health care data from a setting where patients can
experience three events: **Censoring** (0), **Death** (1), and **Disease**
(2). The `beta_X_Y` arguments control how X affects Y — a positive value
means a higher value of X increases the intensity of Y, and a negative value
decreases it. See `?simDisease()` for the full list of arguments.

```{r}
#| label: simdisease

data <- simDisease(
  N = 100,
  cens = 1,
  eta = c(0.1, 0.3, 0.1),
  nu = c(1.1, 1.3, 1.1),
  beta_L0_L = 1,
  beta_A0_L = -1.1,
  beta_L_D = 1,
  beta_L0_D = 0
)

plotEventData(data, title = "Disease data")
```

## Health Care Data with `simTreatment` and `simDropIn`

`simTreatment()` extends the disease setting with a treatment process: it
simulates four event types representing censoring (0), death (1), treatment
(2), and covariate change (3), again controlled via `beta_X_Y` arguments:

```{r}
#| label: simtreatment

data <- simTreatment(
  N = 100,
  beta_L_A = 1,
  beta_L_D = 1,
  beta_A_D = -1,
  beta_A_L = -0.5,
  beta_L0_A = 1,
  eta = rep(0.1, 4),
  nu = rep(1.1, 4),
  followup = 5,
  cens = 1,
  op = 1
)

plotEventData(data, title = "Treatment setting")
```

`simDropIn()` is a related wrapper for a setting with 4 or 5 events —
censoring, death, drop-in initiation, covariate change, and optionally
treatment — again configured through `beta_X_Y` arguments; see
`?simDropIn()` for details.

## Time-Varying Effects with `simEventTV()`

`simEventTV()` adapts `simEventData()` to allow effects that change at a fixed
time point: after time `t_prime`, the matrix `tv_eff` is added to `beta`.
This flexibility comes at the cost of a slightly slower simulation procedure,
so use `simEventData()` instead when effects are constant over time.

```{r}
#| label: simeventtv

set.seed(6258)
eta <- rep(0.1, 4)
term_deltas <- c(0, 1)
tv_eff <- matrix(0.5, ncol = 4, nrow = 6)
beta <- matrix(nrow = 6, ncol = 4, 0.5)
t_prime <- 1

data <- simEventTV(
  N = 100,
  t_prime = t_prime,
  tv_eff = tv_eff,
  eta = eta,
  term_deltas = term_deltas,
  beta = beta,
  lower = 10^(-15),
  upper = 10^2,
  max_events = 30
)

plotEventData(data)
```

## Intensities Depending on Time Since Last Event with `simEventDataTdPhi()`

Sometimes the intensity of an event should depend not just on covariates and
event counts, but on the time elapsed since the most recent occurrence of
that event. Assume the event-specific intensity has the form
$$
  \lambda^x(t \mid \mathcal F_t)
  = R^x(t \mid \mathcal F_t) \,\eta^x \nu^x t^{\nu^x-1} \,\phi^x(t \mid \mathcal F_t),
$$
where
$$
  \phi^x(t \mid \mathcal F_t)
  = \exp\left( L^\top \beta_1 + (T^* - t)\beta_2 \right),
$$
and $T^*$ denotes the most recent event time. Simulation from this setting
is done using `simEventDataTdPhi()`, which takes the additional argument
`beta2` controlling the exponential decay/growth in $\phi^x$:

```{r}
#| label: simeventdatatdphi

beta <- matrix(rnorm(4 * 6, 0, 0.1), ncol = 4)
data <- simEventDataTdPhi(
  N = 100,
  beta2 = c(0.01, 2, 0, 0.1),
  beta = beta,
  upper = 10^20,
  max_iter = 10^3
)
plotEventData(data)
```

# Simulating New Data from a Fitted Model

Besides simulating from user-specified parameters, simevent can simulate
new data from a model already fitted to observed data. `simEventCox()` does
this for Cox proportional hazards models, and `simEventObj()` for any model
equipped with a `predict2` method.

Say we have some observed data:

```{r}
#| label: observed-data

set.seed(373)
beta <- matrix(c(0.5, -1, -0.5, 0.5, 0, 0.5), ncol = 3, nrow = 2)
data <- simCRdata(N = 1000, beta = beta)
```

## Using `simEventCox()`

Fit Cox models for the events of interest, then pass the fitted models to `simEventCox()`:

```{r}
#| label: simeventcox

cox1 <- coxph(Surv(Time, Delta == 1) ~ L0 + A0, data = data)
cox2 <- coxph(Surv(Time, Delta == 2) ~ L0 + A0, data = data)

cox_fits <- list("Process 1" = cox1, "Process 2" = cox2)
old_vars <- data[, c(4, 5)]
new_data <- simEventCox(100, cox_fits, old_vars = old_vars)
head(new_data)
```

New baseline covariates are simulated by resampling rows of `old_vars`.

## Using `simEventObj()`

`simEventObj()` works with any fitted model, as long as you equip it with a
`predict2()` method returning the times and cumulative hazard for new
covariate values. For example, with a Cox model:

```{r}
#| label: simeventobj

cox_fit <- coxph(Surv(Time, Delta == 1) ~ L0 + A0, data = data)

predict2 <- function(obj, ...) UseMethod("predict2")
predict2.coxph <- function(obj, sim_data, ...) {
  preds <- survfit(obj, newdata = sim_data)
  # preds$cumhaz is a times x individuals matrix; reshape to individuals x times x events
  chf <- array(t(preds$cumhaz), dim = c(nrow(sim_data), length(preds$time), 1))
  list(time = preds$time, chf = chf)
}

old_vars <- data[, c("L0", "A0")]
new_data <- simEventObj(100, cox_fit, old_vars = old_vars)
head(new_data)
```

# Interventions

`alphaSim()` and `intEffectAlpha()` simulate data under an intervention that
multiplies the shape parameter of a chosen process by a factor `alpha`, in
one of the predefined settings (`"Disease"`, `"Drop In"`, `"Treatment"`, or `"Statin"`).

Say we want to intervene in the Drop In setting by scaling the intensity of
the drop-in process. Data without intervention is obtained by scaling with
the factor 1:

```{r}
#| label: alphasim

alphaSim(N = 10^3, alpha = 1, tau = 5, setting = "Drop In")
```

By default, `alphaSim()` returns the proportion of individuals who experience
death, and the proportion who experience Drop In, by a specified time
$\tau$, in group `A0 = a0`. Setting `years_lost = TRUE` instead returns the
number of years lost to death and Drop In before $\tau$; setting
`return_data = TRUE` returns the simulated data itself.

`intEffectAlpha()` performs the same kind of intervention and additionally
plots a sample of the simulated event data:

```{r}
#| label: int-effect-alpha

intEffectAlpha(
  N = 1000,
  alpha = 0.7,
  tau = 5,
  years_lost = TRUE,
  a0 = 1,
  plot = TRUE,
  setting = "Drop In"
)
```
